Baseline centering problem

### 2018-12-19 paper/tangotd.tex

The Newton method for matrix inversion is a classical tool to compute
$\Delta^{-1}$. It consists in iterating
\begin{equation}
A\gets 2A-A\Delta A
\end{equation}
It converges extremely fast if initialized reasonably close to $A^{-1}$ (namely if the spectral
radius of $\Id-AX$ is less than $1$), but it can diverge if badly
initialized. A suitable initialization is $A=\eps\transp{\Delta}$ for
small enough $\eps$. For stochastic estimates $\tilde \Delta$ of $\Delta$ (as we have
here through transitions), the stochastic update version with learning
rate $\eta$ is
\begin{equation}
A\gets A+\eta A-\eta A\tilde \Delta A
\end{equation}

@@

Here we do not have access to $\Delta$, but have access to transitions
$s\to s'$, which are samples of
$\Delta$, namely, $\Delta=\sum_s\E_{s'\sim P(s'|s)} \1_s\transp{(\1_s-\gamma
\1_{s'})}$. This corresponds to a stochastic version of the Newton
method, with learning rate $\eta\ll 1$,
\begin{equation}
A\gets A+\eta A-\eta A \E_{ss'} \frac{1}{\rho(s)}\1_s\transp{(\1_s-\gamma \1_{s'})} A
\end{equation}
Indeed, in expectation over $s\to s'$ this corresponds to $A\gets A+\eta
A-\eta A \Delta A$, which converges to $A=\Delta^{-1}$. This
also corresponds to the rank-one update formyla for matrix inversion,
$(D+\eta u\transp{v})^{-1}=D^{-1}-\eta
D^{-1}u\transp{v}D^{-1}+O(\eta^2)$. Namely,
if we update our estimate of $\Delta$ by
$\Delta\gets (1-\eta)\Delta+\eta \frac{1}{\rho(s)} \1_s\transp{(\1_s-\gamma
\1_{s'})}$ when a transition $s\to s'$ is observed, then
$\Delta^{-1}$ gets updated as above.

With $A=\green\cdot \E_{\rho} f$ as above, everything is fully samplable
(thanks to the $\E_\rho$ which cancels the $1/\rho$ factors), see below.
It is not clear to me how to mix the left, right and middle Bellman
equations, but the correspondence with path combinatorics and the Newton
method suggests this
might be an important point for the navigation problem.

Namely, set
\begin{equation}
A=\transp{\green}F, \qquad \green=\sum_s \green(s)\transp{\1_s},\qquad F=\E_{s\sim
\rho}
f(s)\transp{\1_s}
\end{equation}
which is the same representation as above, namely $A=\Delta^{-1}$ is
equivalent to $\green(s)\cdot \E_{s'\sim
\rho} \psi(s')f(s')=\transp{\1_s}\Delta^{-1}\psi$ for all $s$ and test
functions $\psi$. Then the Newton equation for $A$ reads
\begin{equation}
\transp{\green}F\gets (1+\eta)\transp{\green}F-\eta \transp{\green}F \E_{ss'} \frac{1}{\rho(s)}\1_s\transp{(\1_s-\gamma
\1_{s'})} \transp{\green}F
\end{equation}
and since $F\1_s=\rho(s)f(s)$ and
$\transp{\1_s}\transp{\green}=\transp{\green(s)}$, this simplifies to
\begin{equation}
\transp{\green}F\gets (1+\eta)\transp{\green}F-\eta
\transp{\green} \E_{ss'} f(s)\transp{(\green(s)-\gamma
\green(s'))}F
\end{equation}
and we propose to update separately for $\green$ given $f$ and $f$ given
$\green$: the advantage being that for fixed $f$, we can obtain the
update above by updating $\green$ via
\begin{equation}
\transp{\green}\gets (1+\eta)\transp{\green}-\eta
\transp{\green} \E_{ss'} f(s)\transp{(\green(s)-\gamma
\green(s'))}
\end{equation}
and likewise for $f$ given $\green$,
\begin{equation}
F\gets (1+\eta)F-\eta
\E_{ss'} f(s)\transp{(\green(s)-\gamma
\green(s'))}F
\end{equation}
both of which can be implemented via gradient descent on the parameters
of $\green$ and $f$. There is no need to sample on test functions.

>>>>

The Newton method for matrix inversion is a classical tool to compute
$\Delta^{-1}$. It consists in iterating
\begin{equation}
Z\gets 2Z-Z\Delta Z
\end{equation}
It converges extremely fast if initialized reasonably close to $Z^{-1}$ (namely if the spectral
radius of $\Id-ZX$ is less than $1$), but it can diverge if badly
initialized. A suitable initialization is $Z=\eps\transp{\Delta}$ for
small enough $\eps$. For stochastic estimates $\tilde \Delta$ of $\Delta$ (as we have
here through transitions), the stochastic update version with learning
rate $\eta$ is
\begin{equation}
Z\gets Z+\eta Z-\eta Z\tilde \Delta Z
\end{equation}

@@

Here we do not have access to $\Delta$, but have access to transitions
$s\to s'$, which are samples of
$\Delta$, namely, $\Delta=\sum_s\E_{s'\sim P(s'|s)} \1_s\transp{(\1_s-\gamma
\1_{s'})}$. This corresponds to a stochastic version of the Newton
method, with learning rate $\eta\ll 1$,
\begin{equation}
Z\gets Z+\eta Z-\eta Z \E_{ss'} \frac{1}{\rho(s)}\1_s\transp{(\1_s-\gamma \1_{s'})} Z
\end{equation}
Indeed, in expectation over $s\to s'$ this corresponds to $Z\gets Z+\eta
Z-\eta Z \Delta Z$, which converges to $Z=\Delta^{-1}$. This
also corresponds to the rank-one update formyla for matrix inversion,
$(D+\eta u\transp{v})^{-1}=D^{-1}-\eta
D^{-1}u\transp{v}D^{-1}+O(\eta^2)$. Namely,
if we update our estimate of $\Delta$ by
$\Delta\gets (1-\eta)\Delta+\eta \frac{1}{\rho(s)} \1_s\transp{(\1_s-\gamma
\1_{s'})}$ when a transition $s\to s'$ is observed, then
$\Delta^{-1}$ gets updated as above.

With $Z=\green\cdot \E_{\rho} f$ as above, everything is fully samplable
(thanks to the $\E_\rho$ which cancels the $1/\rho$ factors), see below.
It is not clear to me how to mix the left, right and middle Bellman
equations, but the correspondence with path combinatorics and the Newton
method suggests this
might be an important point for the navigation problem.

Namely, set
\begin{equation}
Z=\transp{\green}F, \qquad \green=\sum_s \green(s)\transp{\1_s},\qquad F=\E_{s\sim
\rho}
f(s)\transp{\1_s}
\end{equation}
which is the same representation as above, namely $Z=\Delta^{-1}$ is
equivalent to $\green(s)\cdot \E_{s'\sim
\rho} \psi(s')f(s')=\transp{\1_s}\Delta^{-1}\psi$ for all $s$ and test
functions $\psi$. Then the Newton equation for $Z$ reads
\begin{equation}
\transp{\green}F\gets (1+\eta)\transp{\green}F-\eta \transp{\green}F \E_{ss'} \frac{1}{\rho(s)}\1_s\transp{(\1_s-\gamma
\1_{s'})} \transp{\green}F
\end{equation}
and since $F\1_s=\rho(s)f(s)$ and
$\transp{\1_s}\transp{\green}=\transp{\green(s)}$, this simplifies to
\begin{equation}
\transp{\green}F\gets (1+\eta)\transp{\green}F-\eta
\transp{\green} \E_{ss'} f(s)\transp{(\green(s)-\gamma
\green(s'))}F
\end{equation}
and we propose to update separately for $\green$ given $f$ and $f$ given
$\green$: the advantage being that for fixed $f$, we can obtain the
update above by updating $\green$ via
\begin{equation}
\transp{\green}\gets (1+\eta)\transp{\green}-\eta
\transp{\green} \E_{ss'} f(s)\transp{(\green(s)-\gamma
\green(s'))}
\end{equation}
and likewise for $f$ given $\green$,
\begin{equation}
F\gets (1+\eta)F-\eta
\E_{ss'} f(s)\transp{(\green(s)-\gamma
\green(s'))}F
\end{equation}
both of which can be implemented via gradient descent on the parameters
of $\green$ and $f$. (This double the learning rate $\eta$.)
There is no need to sample on test functions.

(Equivalent to solving for $ \Id=\E_{ss'} f(s)\transp{(\green(s)-\gamma
\green(s'))}$, namely, $\Id=F\Delta\transp{\green}$ in the representation space $E$, but via a different gradient
descent.)

The Newton version should not be used alone. Indeed, with a matrix $Z$ of
rank $r$, it just settles on some $r$-dimensional subspace and inverts 
$\Delta^{-1}$ on that subspace. The subspace in question does not
change at all along optimization. (This is because the equations on $F$
and $\green$ leave $\Ker F$ and $\Img \transp{\green}$ invariant; more
precisely, $\Delta Z=\Id$ on $\Img \transp{\green}$ and $\Img(\Delta
Z-\Id)\subset\Ker F$)
From a purely algebraic point of view,
all these $r$-dimensional subspaces are equally important: there are as
many quasi-inverses as such subspaces, and the subspace is fixed at the
initialization of $F$ and $\green$. (For instance, if $r=1$ one can
choose any vector for $\transp{\green}$, and choose for $F$ any linear
form that evaluates to $1$ on $\Delta \transp{\green}$).

On the other hand the left and right Bellman equations prioritize
inverting $\Delta$ on test functions $\psi$. So this introduces
(implicitly) a quantitative criterion that breaks symmetry between
possible quasi-inverses.

Intuitively, the middle Bellman equation does not create new starting
points or endpoints on paths, only works in the middle. (This corresponds
to the fixed image and kernel.) Meanwhile the left Bellman equation
creates starting points and the right Bellman equation creates endpoints
when observing a new transition $s\to s'$. This introduces spatial
smoothness over the image and kernel.

I still don't know how to rightly combine these three equations.
